Constructing a universe for the setoid model

### 2021-01-22 paper.tex
??

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{bbm}
\usepackage{mathtools}
%\usepackage[bw]{agda}

@@

% for placeholders
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}

% for definitions
\newcommand{\defeq}{\mathrel{\vcentcolon\equiv}}

@@

\author{Thorsten Altenkirch\inst{1}
  \thanks{Supported by USAF grant FA9550-16-1-0029.}
  \orcidID{0000-0002-6582-5025} \and
  Simon Boulier\inst{2}\thanks{Supported by ERC Starting Grant CoqHoTT 637339.} \and
  Ambrus Kaposi\inst{3}\orcidID{0000-0001-9897-8936}
  \thanks{Supported by the Bolyai Fellowship of the Hungarian Academy
    of Sciences (BO/00659/19/3) and by the ``Application Domain
    Specific Highly Reliable IT Solutions'' project that has been
    implemented with support from the National Research, Development
    and Innovation Fund of Hungary, financed under the Thematic
    Excellence Programme TKP2020-NKA-06 funding scheme.}\and
  Christian Sattler\inst{4}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.} \and
  Filippo Sestini\inst{1}\orcidID{0000-0002-8701-5613}}

@@

One of the main appeals of Altenkirch's setoid model is that it can justify
several useful extensionality principles while being defined in a small
intensional metatheory.
%
We tried to stay true to this idea when figuring out the necessary
metatheoretical tools for the universe construction in this paper. In
particular, we wanted to avoid having to assume strong definition schemas that
go beyond inductive families.
$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type theory.

We have sorts $\Type_i$ of types and $\mProp_i$ of strict propositions for $i \in \{0, 1\}$.
Here, $i = 0$ means ``small'' (and we will omit the subscript) and $i = 1$ means ``large''.
We have implicit lifting from $i = 0$ to $i = 1$, but do not assume type formers are preserved.
$\Type_1$ has universes for $\Type$ and $\mProp$.
We do not distinguish notationally between universes and sorts.
We continue to describe only the case $i = 0$; everything introduced has an analogue at level $i = 1$.
Propositions lift to types via $\textsf{Lift} : \mProp \to \mType$, with constructor
$\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
$\textsf{unlift} : \{P : \mProp \} \to \textsf{Lift}\ P \to P$.

We have standard type formers $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$ in $\Type$.
$\Sigma$-types are defined negatively by pairing $\blank,\blank$ and projections $\pi_1, \pi_2$.
We have definitional $\eta$-rules for $\Pi$-, $\Sigma$-, $\mathbf{1}$-types.
We also require indexed W-types, both in $\mType$ and $\mProp$:
$W_\Box : (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
The elimination principle of $W_{\mProp}$ only allows defining functions into elements of $\mProp$.
%
From $W_{\mProp}$ we can define propositional truncation $\|\blank\| : \mType \to \mProp$, with constructor
  $|\blank| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\blank\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
\vspace{-0.2em}
\begin{align*}
  \mathbf{1}_\mProp & \defeq \| \mathbf{1} \| \\
  \Pi_\mProp\ P\ Q & \defeq \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_\mProp\ P\ Q & \defeq \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
\end{align*}
We assume that we have
$\mathbf{0}_\mProp:\mProp$ together with $\mathsf{exfalso}_\mProp :\{A:\mType \} \to \mathbf{0}_\mProp \to A$.

@@

For convenience, we slightly deviate from $\mlttp$ both in the paper and in the
formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
types instead of deriving them from truncation. We operate under the assumption
that everything can be equivalently carried out in $\mlttp$, although we have not
fully checked all the necessary details.

@@

We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), function extensionality and a universe of strict
propositions with propositional extensionality \cite{setoid99}. Note that
we do not need identity types or inductive types (W-types) for this.

@@

Notably, equality types in SeTT compute definitionally on concrete type
formers. In particular, they compute to their obvious intended meaning, so that
an equality of pairs is a pair of equalities, an equality of functions is a map
of equalities, and so on. From this, we get definitional versions of function
and propositional extensionality.
%% %
%% \begin{align*}
%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)\ldots \\
%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = \ldots \\
%%   & \Prop^\sim\ p\ P\ Q = \ldots
%% \end{align*}

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.
\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 \defeq A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t \defeq
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

@@

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such an elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it is
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such a universe eliminator.
%
The question arises because our final encoding of the setoid universe only
supports a weak form of elimination, for reasons that are explained in
Section~\ref{inductive-universe}. Although not currently needed, a stronger
eliminator might be necessary to justify universe induction. This problem should
not arise in the other encodings of the setoid universe (as given in
Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$. \\
\vspace{-0.2em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

Induction-recursion is arguably a nice and natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $A : \mType$ with $A \to I$) \cite{malatasta13smallir}.

@@

$\U$ and $\El$ can be given by $\U \defeq \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El \defeq \pi_1$.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \blank\sim_\setoidU\blank : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \blank\vdash\blank\sim_\El\blank : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | \defeq \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim \defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty \defeq \reflu \\
  & \ldots \\
  & \textsf{coe}\ \Uty \defeq \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty \defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | \defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim \defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) \defeq \reflel \\
  & \ldots \\
  & \textsf{coe}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) \defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

In the next section we transform our current inductive-recursive definition to
one that does not use induction-recursion.
%
The way this is done is inspired by the well-known trick to eliminate
induction-recursion described in Section~\ref{ir-universes}, but modified in a
novel way to account for the presence of $\mProp$-valued types. To our
knowledge, this is the first time this reduction method is applied to an
inductive-recursive type of this kind.

@@

Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
extension $\blank,\blank$. Types in \textsf{Ty} are either the base type $\iota$ or
$\Pi$ types.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
  & \blank,\blank && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
\end{alignat*}

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
\vspace{-0.2em}
\begin{minipage}{0.4\textwidth}
\begin{alignat*}{10}
  & \bullet_0 && : \textsf{Con}_0 \\
  & \blank,_0\blank && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
  & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{alignat*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
  & \blank,_1\blank && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 && : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 && : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}
\\

>>>>

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{bbm}
\usepackage{marvosym}
%\usepackage[bw]{agda}

@@

@@

\author{Thorsten Altenkirch\inst{1}
  \thanks{Supported by USAF grant FA9550-16-1-0029.}
  \orcidID{0000-0002-6582-5025}\Letter \and
  Simon Boulier\inst{2}\thanks{Supported by ERC Starting Grant CoqHoTT 637339.} \and
  Ambrus Kaposi\inst{3}\orcidID{0000-0001-9897-8936}
  \thanks{Supported by the Bolyai Fellowship of the Hungarian Academy
    of Sciences (BO/00659/19/3) and by the ``Application Domain
    Specific Highly Reliable IT Solutions'' project that has been
    implemented with support from the National Research, Development
    and Innovation Fund of Hungary, financed under the Thematic
    Excellence Programme TKP2020-NKA-06 funding scheme.}\and
  Christian Sattler\inst{4}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.} \and
  Filippo Sestini\inst{1}\orcidID{0000-0002-8701-5613}}

@@

One of the main appeals of Altenkirch's setoid model is that it can justify
several useful extensionality principles while being defined in a small
intensional metatheory.
%
We tried to stay true to this idea when figuring out the necessary
metatheoretical tools for the universe construction in this paper. In
particular, we wanted to avoid having to assume strong definition schemas that
go beyond inductive families.

$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type
theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
types via a lifting $\textsf{Lift} : \mProp \to \mType$, with constructor
$\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
$\textsf{unlift} : \{P : \mProp \} \to \textsf{Lift}\ P \to P$.

We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
: (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
The elimination principle of $W_\mProp$ only allows defining functions into $\mProp$s.
%
From $W_\mProp$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
\vspace{-0.2em}
\begin{align*}
  \mathbf{1}_\mProp & :\equiv \| \mathbf{1} \| \\
  \Pi_\mProp\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_\mProp\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
\end{align*}
We assume that we have
$\mathbf{0}_\mProp:\mProp$ together with $\mathsf{exfalso}_\mProp :\{A:\mType \} \to \mathbf{0}_\mProp \to A$.

@@

For convenience, we slightly deviate from $\mlttp$ both in the paper and in the
formalization, for instance by relying on pattern matching instead of
eliminators, and using primitive versions of $\mProp$-valued $\Pi$ and $\Sigma$
types instead of deriving them from truncation. We operate under the assumption
that everything can be equivalently carried out in $\mlttp$, although we haven't
fully checked all the necessary details.

@@

We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), function extensionality and a universe of strict
propositions with propositional extensionality \cite{setoid99}. Note that
we don't need identity types or inductive types (W-types) for this.

@@

Notably, equality types in SeTT compute definitionally on concrete type
formers. In particular, they compute to their obvious intended meaning, so that
an equality of pairs is a pair of equalities, an equality of functions is a map
of equalities, and so on. From this, we get definitional versions of function
and propositional extensionality.
%% %
%% \begin{align*}
%%   & (\Pi(a : A)\ B)^\sim\ p\ f_0\ f_1 = \Pi(a_0\,a_1 : A)... \\
%%   & (\Sigma\ A\ B)^\sim\ p\ (a_0, b_0)\ (a_1, b_1) = ... \\
%%   & \Prop^\sim\ p\ P\ Q = ...
%% \end{align*}

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.
\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t :\equiv
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

@@

Other design choices on the setoid universe are less essential, but still
require careful consideration. For instance, one question is whether the setoid
universe should support universe induction, thus exposing the inductive
structure of the codes. Such an elimination principle is known to be inconsistent
with univalence, although this is not an issue in our case; nevertheless it's
not immediately clear if the elimination principle can be justified by the
semantics, that is, if our encoding of the setoid universe in the model allows
to define such a universe eliminator.
%
The question arises because our final encoding of the setoid universe only
supports a weak form of elimination, for reasons that are explained in
Section~\ref{inductive-universe}. Although not currently needed, a stronger
eliminator might be necessary to justify universe induction. This problem should
not arise in the other encodings of the setoid universe (as given in
Section~\ref{ir-setoid-universe} and Section~\ref{ii-universe}).

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$. \\
\vspace{-0.2em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

Induction-recursion is arguably a nice and natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$) \cite{malatasta13smallir}.

@@

$\U$ and $\El$ can be given by $\U :\equiv \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
\vspace{-0.2em}
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
\vspace{-0.2em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

In the next section we transform our current inductive-recursive definition to
one that doesn't use induction-recursion.
%
The way this is done is inspired by the well-known trick to eliminate
induction-recursion described in Section~\ref{ir-universes}, but modified in a
novel way to account for the presence of $\mProp$-valued types. To our
knowledge, this is the first time this reduction method is applied to an
inductive-recursive type of this kind.

@@

Contexts in \textsf{Con} are formed out of empty contexts $\bullet$ and context
extension $\_,\_$. Types in \textsf{Ty} are either the base type $\iota$ or
$\Pi$ types.
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && : \textsf{Con}                                                          &&  \iota && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \\
  & \_,\_ && : (\GG : \textsf{Con}) \to \textsf{Ty}\ \GG \to \textsf{Con} \hspace{3em} &&  \Pi && : \{\GG : \textsf{Con}\} (A : \textsf{Ty}\ \GG) \to \textsf{Ty}\ (\GG , A) \to \textsf{Ty}\ \GG
\end{alignat*}

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
\vspace{-0.2em}
\begin{minipage}{0.4\textwidth}
\begin{alignat*}{10}
  & \bullet_0 && : \textsf{Con}_0 \\
  & \_,_0\_ && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  & \iota_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \\
  & \Pi_0 && : \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{alignat*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  & \bullet_1 && : \textsf{Con}_1\ \bullet_0 \\
  & \_,_1\_ && : \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & && \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  & \iota_1 && : \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  & \Pi_1 && : \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & && \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}
\\

We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
\Sigma\ (\GG_0 : \textsf{Con}_0)\ (\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
:\equiv \Sigma\ (A_0 :
\textsf{Ty}_0)\ (\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
is straightforward:
\vspace{-0.2em}
\begin{alignat*}{10}
  & \bullet && :\equiv (\bullet_0 , \bullet_1) \\
  & (\GG_0 , \GG_1) , (A_0 , A_1) && :\equiv ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
  & \iota\ (\GG_0 , \GG_1) && :\equiv (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && :\equiv
    (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
\end{alignat*}
